---
title: "AI & the Future of Work: Skills Demand Insights"
author: "Amenah Altamimi, Bryant Ha, Shaza Mumtaz" 
format:
  revealjs:
    slide-number: true
    chalkboard: true
    footer: "AI Job Market Analysis"
    transition: fade # Added transition effect
    transition-speed: default 
    theme: solarized
    css: theme.scss
    logo: "images/ccidm-cpp_logo.png"
---

```{r setup, include=FALSE}
# Load necessary libraries and prepare data for RO3
library(tidyverse)
library(gridExtra)

df <- read_csv("ai_job_market_insights.csv")
ro3_data <- df %>%
  filter(
    !is.na(Required_Skills),
    !is.na(Remote_Friendly),
    !is.na(Industry),
    !is.na(Salary_USD)
  ) %>%
  separate_rows(Required_Skills, sep = ",\\\\s*") %>%
  mutate(
    Remote = if_else(Remote_Friendly == "Yes", "Remote", "In-Person"),
    Salary_Bin = if_else(
      Salary_USD > median(Salary_USD, na.rm = TRUE),
      "High Salary", "Low Salary"
    )
  )

# Determine top 5 skills overall
top5 <- ro3_data %>%
  count(Required_Skills, name = "Overall") %>%
  arrange(desc(Overall)) %>%
  slice_head(n = 5) %>%
  pull(Required_Skills)

ro3_top5 <- ro3_data %>% filter(Required_Skills %in% top5)

# Create summary tables for each factor
remote_counts   <- ro3_top5 %>% count(Required_Skills, Remote,   name = "Count")
industry_counts <- ro3_top5 %>% count(Required_Skills, Industry, name = "Count")
salary_counts   <- ro3_top5 %>% count(Required_Skills, Salary_Bin, name = "Count")
```

## Project Overview
**AI & the Future of Work** examines how AI adoption reshapes jobs and skills globally. Using a dataset of 500+ job listings from Singapore, Dubai, and Berlin across tech, marketing, and retail, we explore:

- **RO1:** AI adoption’s link to job growth/decline
- **RO2:** Roles most impacted by AI and drivers of demand
- **RO3:** In-demand skills in an AI-driven market and the effect of remote vs. in-person work

---

## Research Objective 3: Top Skills Analysis
- **IV:** Job Title (skill proxy)
- **DV:** Required_Skills frequency
- **Factors:** Work Mode, Industry, Salary Tier
- **Goal:** Reveal how these dimensions shape the top five skills

---

## Skills by Work Mode

```{r remote-vs, echo=FALSE, fig.height=4}
ggplot(remote_counts, aes(
  x = reorder(Required_Skills, Count),
  y = Count,
  fill = Remote
)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  labs(
    title = "Top 5 Skills: Remote vs. In-Person",
    x = NULL, y = "Count", fill = "Work Mode"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title       = element_text(hjust = 0.5, face = "bold"),
    panel.grid.major = element_line(linetype = "dotted"),
    panel.grid.minor = element_blank()
  )
```

**Key Insights:**
- **Communication** excels In-Person
- **Python** leads Remote
- Informs targeted upskilling by mode

---

## Industry & Salary Effects

```{r industry-salary, echo=FALSE, fig.height=6}
p1 <- ggplot(industry_counts, aes(
  x = reorder(Required_Skills, Count),
  y = Count,
  fill = Industry
)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  labs(
    title = "Top 5 Skills by Industry",
    x = NULL, y = "Count", fill = "Industry"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title       = element_text(hjust = 0.5, face = "bold"),
    panel.grid.major = element_line(linetype = "dotted"),
    panel.grid.minor = element_blank()
  )

p2 <- ggplot(salary_counts, aes(
  x = reorder(Required_Skills, Count),
  y = Count,
  fill = Salary_Bin
)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  labs(
    title = "Top 5 Skills by Salary Tier",
    x = NULL, y = "Count", fill = "Salary Level"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title       = element_text(hjust = 0.5, face = "bold"),
    panel.grid.major = element_line(linetype = "dotted"),
    panel.grid.minor = element_blank()
  )

grid.arrange(p1, p2, ncol = 1)
```

**Industry Trends:**
- Tech → _Machine Learning_, _Data Analysis_
- Marketing/Retail → _SEO_, _Creative Suite_

**Salary Trends:**
- High-pay → _Cloud Architecture_, _AI Ethics_
- Low-pay → _MS Office_, _Basic SQL_