---
title: "AI & the Future of Work: Skills Demand Insights"
author: "Amenah Altamimi, Bryant Ha, Shaza Mumtaz"
institute: “GBA 5910, Cal Poly Pomona"
date: “05-12-2025
format: 
  revealjs:
    theme: serif
    width: 1600
    height: 900
    footer: “Final Project"
    transition: fade
    transition-speed: slow
    incremental: false
    toc: true
    toc-depth: 1
    slide-number: true
    scrollable: true
    smaller: true
    code-fold: false
    code-overflow: wrap
    number-sections: false
    number-depth: 5
    embed-resources: true
    css: styles.css
editor: visual
execute: 
  echo: true
  freeze: auto
---

<<<<<<< HEAD
=======
```{r setup, include = FALSE}
library(tidyverse)
library(tidyr)    
library(ggthemes)
library(knitr)
library(MASS)
```

>>>>>>> e7823b152863fa1f8efc52cfb455b7535f248fbd
## Table of Contents

-   Introduction
-   Problem Definition
-   Research Objectives
-   Importance of the Research
-   Population & Sample Collection
-   R02: Identify which job roles are most affected by AI
-   RO3: Skills Demand Insights

## Introduction

Welcome to our analysis on how AI adoption is transforming the global
job market. This presentation focuses on identifying emerging skill
demands and the influence of remote versus in-person work environments.

## Problem Definition

Rapid AI integration is reshaping job roles and required skill sets.
Organizations face challenges in workforce planning, upskilling, and
aligning recruitment with evolving market needs.

## Research Objectives

1.  **RO1:** Assess the impact of AI adoption on overall job growth or
    decline.
2.  **RO2:** Identify which job roles are most affected by AI and why.
3.  **RO3:** Determine the most in-demand skills and examine how remote
    vs. in-person work influences skill requirements.

## Importance of the Research

-   **For Businesses:** Guides strategic hiring and training programs.
-   **For Employees:** Clarifies skill development pathways for career
    resilience.
-   **For Policymakers:** Informs education and workforce policies to
    support an AI-driven economy.

## Population & Sampling/Data Collection {transition="zoom"}

All professional job roles (focus: UX/UI designers, project managers,
marketing professionals) globally or in the U.S., across industries
(tech, marketing, etc.), from 2020–2025.

The current convenience sample and pre-compiled data provide a starting
point, but stratified random sampling and enhanced, diverse collection
methods would strengthen reliability and address the topic’s
significance.

------------------------------------------------------------------------

## Data Wrangling {transition="fade"}

```{r}
# Clear environment
rm(list = ls())

# Load libraries
library(readr)
library(dplyr)

# Load data (adjust path as needed)
data <- read_csv("ai_job_market_insights.csv")

# Data wrangling
data_clean <- data %>%
  dplyr::mutate(
    AI_Adoption_Level = factor(AI_Adoption_Level, levels = c("Low", "Medium", "High")),
    Job_Growth_Projection = factor(Job_Growth_Projection, levels = c("Decline", "Stable", "Growth")),
    Salary_USD = as.numeric(Salary_USD)
  ) %>%
  dplyr::filter(!is.na(Salary_USD) & !is.na(AI_Adoption_Level) & !is.na(Job_Growth_Projection))

# Verify
str(data_clean)
```

## Sample Characteristics {transition="fade"}

-   **Population:** All job listings from 2020–2025 across multiple
    industries\
-   **Sample:** Filtered to records with non-missing AI adoption and
    growth projections
-   **Industries covered:** Finance, Technology, Healthcare,
    Manufacturing, etc.

## Research Questions & Hypotheses {transition="fade"}

-   **RQ1 / H1:** Does AI adoption level affect job growth projections?\
-   **RQ2 / H2:** Which job roles are most impacted by AI and why?\
-   **RQ3 / H3:** What skills are most in demand, and how does remote
    vs. in-person work influence those requirements?

## **RQ1 / H1:**

**H1:** Does AI adoption level affect job growth projections?\
**RO1:** Assess the impact of AI adoption on overall job growth or
decline.

## Variables & Their Roles {.scrollable}

::: {style="font-size:0.9em; width:90vw; margin:auto;"}
| Variable          | Type                                   | Role                  |
|-------------------|--------------------------------|---------------------|
| AI_Adoption_Level | Ordered factor (Low, Medium, High)     | Independent Variable  |
| Growth_Score      | Numeric (-1, 0, +1)                    | Dependent Variable    |
| Required_Skills   | Categorical list                       | Dependent (frequency) |
| Remote            | Categorical (Remote / In-Person)       | Independent Variable  |
| Industry          | Categorical                            | Independent Variable  |
| Salary_Bin        | Categorical (High Salary / Low Salary) | Independent Variable  |
:::

## Operationalization of the Scale {transition="zoom"}

-   **AI_Adoption_Level:** Mapped to 1–3 (Low=1, Medium=2, High=3)\
-   **Growth_Score:** Decline = –1; Stable = 0; Growth = +1\
-   **Required_Skills:** Split multi-skill entries into single rows,
    counted frequencies\
-   **Remote:** Derived from `Remote_Friendly` (“Yes” → Remote; “No” →
    In-Person)\
-   **Salary_Bin:** Above vs. below median `Salary_USD`

## Data Wrangling {transition="fade"}

1.  Loaded raw CSV via `read_csv()`\
2.  Filtered out missing values in key fields\
3.  Converted relevant columns to factors and created `Growth_Score`\
4.  Separated `Required_Skills` into individual records

```{r data_prep, include=FALSE}
# 1. Data Preparation ------------------------------------------------------
# Load and clean data
df <- read_csv("ai_job_market_insights.csv")

ro1_data <- df %>%
  filter(!is.na(AI_Adoption_Level), !is.na(Job_Growth_Projection)) %>%
  mutate(
    AI_Adoption = factor(AI_Adoption_Level, 
                         levels = c("Low", "Medium", "High"),
                         ordered = TRUE),
    Growth_Score = case_when(
      Job_Growth_Projection == "Decline" ~ -1,
      Job_Growth_Projection == "Stable" ~ 0,
      Job_Growth_Projection == "Growth" ~ 1
    )
  )
```

```{r desc_stats, include=FALSE}

# 2. Descriptive Statistics ------------------------------------------------
desc_stats <- ro1_data %>%
  count(AI_Adoption, Job_Growth_Projection) %>%
  pivot_wider(names_from = Job_Growth_Projection, values_from = n)

kable(desc_stats, caption = "Job Counts by AI Adoption Level")
```
## Data Visualization - Job Growth Proportions {transition="fade"}

```{r growth-proportion, fig.width=8, fig.height=6}
ggplot(ro1_data, aes(x = AI_Adoption, fill = Job_Growth_Projection)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Job Growth Proportion by AI Adoption Level",
    x = "AI Adoption Level",
    y = "Proportion",
    fill = "Growth Status"
  ) +
  theme_minimal()
```

## Data Visualization - Growth Scores {transition="fade"}

```{r growth-boxplot, fig.width=8, fig.height=6}
ggplot(ro1_data, aes(x = AI_Adoption, y = Growth_Score, fill = AI_Adoption)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Job Growth Score by AI Adoption Level",
    x = "AI Adoption Level",
    y = "Growth Score (Decline=-1, Stable=0, Growth=1)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Data Visualization - Industry Facets {transition="fade"}

```{r industry-facet, fig.width=8, fig.height=6}
ggplot(ro1_data, aes(x = AI_Adoption, fill = Job_Growth_Projection)) +
  geom_bar(position = "dodge") +
  facet_wrap(~Industry, ncol = 2) +
  labs(
    title = "Job Growth by AI Adoption Across Industries",
    x = "AI Adoption Level",
    y = "Count",
    fill = "Growth Status"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Summary Findings {transition="zoom"}

-   No significant association between AI adoption and job growth (χ² p
    \> 0.05)\
-   Remote roles skew technical (Python, Cloud Architecture); in-person
    skew soft skills (Communication, Teamwork)\
-   Transportation & Finance show strongest growth at medium AI
    adoption\
-   **Business takeaway:** Tailor upskilling and hiring to skill demands
    by work mode and industry

## RO2: Roles Most Impacted by AI and Drivers of Demand {transition="zoom"}

**Prompt:**\
We will do some Data Cleaning to prepare, and create demand indicator
and risk scores.

### Overview

::: incremental
-   Investigate job roles’ vulnerability to AI automation.
-   Analyze drivers of demand (e.g., salary, remote work).
-   Use R to quantify automation risk and demand trends.
:::

## Data Preparation: Loading Libraries {transition="fade"}

### Setup

::: incremental
-   Load `tidyverse`, `dplyr`, `ggplot2`, and `readr` for wrangling and
    visualization.
-   Ensure consistent, reproducible analysis.
:::

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readr)
```

## Read,Inspect the data & Conversion {transition="fade"}

```{r}
# Read the CSV file
library(readr)
ai_job <- read_csv("ai_job_market_insights.csv")

# Inspect the data without X11
head(ai_job)        # first few rows
glimpse(ai_job)     # structure summary
summary(ai_job)     # basic stats

# Data Wrangling: convert categorical variables to factors
ai_job <- ai_job %>%
  mutate(
    Job_Title           = as.factor(Job_Title),
    Industry            = as.factor(Industry),
    Company_Size        = Company_Size,        # adjust if needed
    Location            = as.factor(Location),
    AI_Adoption_Level   = as.factor(AI_Adoption_Level),
    Automation_Risk     = as.factor(Automation_Risk),
    Remote_Friendly     = as.factor(Remote_Friendly),
    Job_Growth_Projection = as.factor(Job_Growth_Projection)
  )

# Check for missing data
colSums(is.na(ai_job))

```

## Create a risk scores & Demand Level {transition="fade"}

::: increnental
Based on Demand Level and Automation Risk - **Variables**: Independent:
AI_Adoption_Level (Low=1, Medium=2, High=3). Dependent: Automation_Risk
(Low=1, Medium=2, High=3)
:::

```{r}
# Now we are going to create a risk score (High scores = HIgh risk) based on the Automation_Risk , it helps quantify the risk level for comparison
ai_job <- ai_job %>%
  mutate(Automation_Risk_Score = case_when(Automation_Risk == "Low" ~ 1,
                                     Automation_Risk == "Medium" ~ 2,
                                     Automation_Risk == "High" ~ 3
                                     ))
# Next, we will create a demand indicator based on Job_Growth_Projection , simplifies the growth project into a three tier demand scale 
ai_job <- ai_job %>%
  mutate(Demand_Level = case_when(
    Job_Growth_Projection == "Growth" ~ "High",
    Job_Growth_Projection == "Stable" ~ "Medium",
    Job_Growth_Projection == "Decline" ~ "Low"
  ))
```

## Job Most Impacted by AI {transition="zoom"}

Prompt: Identify which jobs are the most affected by AI

::: incremental
-   **Variables**: Independent: AI_Adoption_Level (Low=1, Medium=2,
    High=3). Dependent: Automation_Risk (Low=1, Medium=2, High=3).
-   **Approach**: Summarize automation risk by job title, focusing on
    high AI adoption.
-   **Findings**: UX Designers and Marketing Specialists face high
    automation risk in high-AI industries (e.g., Technology).
:::

```{r}
ai_impact <- ai_job %>%
  group_by(Job_Title) %>%
  summarise(
    Avg_Automation_Risk = mean(Automation_Risk_Score),
    High_Risk_Count = sum(Automation_Risk == "High"),
    AI_Adoption_High = sum(AI_Adoption_Level == "High")
  ) %>%
  arrange(desc(Avg_Automation_Risk))
print(ai_impact)

#View top jobs with highest automation risk
print(ai_impact)

# Focus on specific roles (e.g., UX Designer, Project Manager, Marketing Specialist)
target_jobs <- c("UX Designer", "Project Manager", "Marketing Specialist")
ai_impact_target <- ai_impact %>% filter(Job_Title %in% target_jobs)
print(ai_impact_target)
```

## Demand Trends {transition="“zoom”"}

Prompt: The Demand Trend.

::: incremental
-   **Variables**: Independent: Salary_USD (continuous). Dependent:
    Demand_Level (High, Medium, Low).
-   **Approach**: Analyze demand by job title, focusing on
    stable/growing roles.
-   **Findings**: Project Manager and AI Researcher show high/medium
    demand.Human-centric skills (e.g., communication) and specialized
    expertise (e.g., machine learning) resist automation.
:::

```{r}
# Analyze demand by job title
demand_analysis <- ai_job %>%
  group_by(Job_Title, Demand_Level) %>%
  summarise(
    Avg_Salary = mean(Salary_USD),
    Remote_Percent = mean(Remote_Friendly == "Yes") * 100,
    Total_Jobs = n()
  ) %>%
  arrange(Job_Title, desc(Demand_Level))

# View demand trends
print(demand_analysis)

# Visualize demand vs. salarynd remote work
ggplot(demand_analysis, aes(x = Avg_Salary, y = Remote_Percent, color = Demand_Level, size = Total_Jobs)) +
  geom_point() +
  labs(title = "Job Demand vs. Salary and Remote Work", x = "Average Salary (USD)", y = "Percentage Remote Friendly") +
  theme_minimal()
```

## Pay and Remote Work Influence {transition="“zoom”"}

Prompt: Explore Influence of Pay and Remote Work.

::: incremental
-   **Variables**: Salary_USD, Remote_Friendly, Demand_Level.
-   **Findings**: Marketing Specialists (\$93K, 60% remote) decline,
    while Project Managers (\$92K, 55% remote) grow, suggesting skills
    drive demand over pay.
:::

```{r}
# Compare salary and remote work by demand
pay_remote_demand <- ai_job %>%
  group_by(Demand_Level) %>%
  summarise(
    Avg_Salary = mean(Salary_USD),
    Remote_Percent = mean(Remote_Friendly == "Yes") * 100,
    Job_Count = n()
  )

print(pay_remote_demand)
```

## Visualization {transition="“fade”"}

Salary by Demand

```{r}
# Visualize
ggplot(pay_remote_demand, aes(x = Demand_Level, y = Avg_Salary, fill = Remote_Percent)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Salary by Demand Level with Remote Work Influence", x = "Demand Level", y = "Average Salary (USD)") +
  theme_minimal() +
  scale_fill_gradient(low = "lightblue", high = "darkblue")
```

## RO2 Conclusions {transition="“zoom”"}

Key Findings ::: incremental - **High-Risk Jobs**: UX Designers and
Marketing Specialists face high automation risk in high-AI industries
(e.g., Technology). - **In-Demand Jobs**: AI Researchers and Project
Managers show stable/growing demand due to specialized and human-centric
skills. - **Pay/Remote Work**: Higher salaries and remote options
correlate with demand, but human judgment is critical. :::

# RQ3 / H3

**Research Question 3 (RQ3):** What skills are most in demand in an
AI-driven job market, and how does remote work versus in-person work
affect these skill requirements?

**Hypothesis 3 (H3):** The prevalence of certain skills differs
significantly between remote and in-person roles, with technical skills
(e.g., Python, Cloud Architecture) being more common in remote positions
and soft skills (e.g., Communication, Teamwork) more common in in-person
roles.

## Variables and Their Roles {.scrollable}

::: {style="font-size:0.8em; width:90vw; margin:auto;"}
| Variable            | Role                    | Description                                                |
|------------------|------------------|-------------------------------------|
| **Required_Skills** | Dependent Variable (DV) | Count of each individual skill extracted from job listings |
| **Remote**          | Independent Variable 1  | Categorical: “Remote” vs. “In-Person”                      |
| **Industry**        | Independent Variable 2  | Categorical: sector of the job (e.g., Tech, Marketing)     |
| **Salary_Bin**      | Independent Variable 3  | Categorical: “High Salary” vs. “Low Salary”                |
:::

# Operationalization of the Scale

-   **Required_Skills:** Each skill listed in `Required_Skills` is split
    into one row per skill; frequency counts measure demand.\
-   **Remote:** Derived from `Remote_Friendly` column; binary factor.\
-   **Industry:** Directly from `Industry` field; nominal categories.\
-   **Salary_Bin:** Binned at the median `Salary_USD`; above = "High
    Salary", below = "Low Salary".

## Data Wrangling {.scrollable}

```{r}
library(tidyverse)

# Load raw data
df <- read_csv("ai_job_market_insights.csv")

# Clean, split skills, and create flags
ro3_data <- df %>%
  filter(
    !is.na(Required_Skills),
    !is.na(Remote_Friendly),
    !is.na(Industry),
    !is.na(Salary_USD)
  ) %>%
  separate_rows(Required_Skills, sep = ",\\s*") %>%
  mutate(
    Remote     = if_else(Remote_Friendly == "Yes", "Remote", "In-Person"),
    Salary_Bin = if_else(
      Salary_USD > median(Salary_USD, na.rm = TRUE),
      "High Salary", 
      "Low Salary"
    )
  )

# Select top 5 most common skills overall
top5_skills <- ro3_data %>%
  count(Required_Skills, name = "Total") %>%
  slice_max(Total, n = 5) %>%
  pull(Required_Skills)

# Subset to top 5
ro3_top5        <- ro3_data %>% filter(Required_Skills %in% top5_skills)
remote_counts   <- ro3_top5 %>% count(Required_Skills, Remote,   name = "Count")
industry_counts <- ro3_top5 %>% count(Required_Skills, Industry, name = "Count")
salary_counts   <- ro3_top5 %>% count(Required_Skills, Salary_Bin, name = "Count")

# Display the full Remote vs In-Person table
knitr::kable(
  remote_counts,
  caption = "Top-5 Skills: Remote vs In-Person",
  align   = c("l","l","r")
)
```

# Data Visualization

## Skills by Work Mode

```{r viz-remote, echo=FALSE, fig.height=4}
ggplot(remote_counts, aes(x = reorder(Required_Skills, Count), y = Count, fill = Remote)) +
  geom_col(position = "dodge") + coord_flip() +
  labs(title = "Top 5 Skills: Remote vs. In-Person", x = NULL, y = "Count") +
  theme_minimal()
```

## Skills Across Industries

```{r viz-industry, echo=FALSE, fig.height=4}
ggplot(industry_counts, aes(x = reorder(Required_Skills, Count), y = Count, fill = Industry)) +
  geom_col(position = "dodge") + coord_flip() +
  labs(title = "Top 5 Skills by Industry", x = NULL, y = "Count") +
  theme_minimal()
```

## Skills by Salary Tier

```{r viz-salary, echo=FALSE, fig.height=4}
ggplot(salary_counts, aes(x = reorder(Required_Skills, Count), y = Count, fill = Salary_Bin)) +
  geom_col(position = "dodge") + coord_flip() +
  labs(title = "Top 5 Skills: High vs. Low Salary", x = NULL, y = "Count") +
  theme_minimal()
```

# Summary Findings

-   **Remote vs. In-Person:** Remote roles demand more technical skills
    (e.g., *Python*, *Cloud Architecture*), whereas in-person roles
    emphasize soft skills (e.g., *Communication*, *Teamwork*).\
-   **Industry Differences:** Tech roles favor *Machine Learning* and
    *Data Analysis*, while Marketing/Retail focus on *SEO* and *Creative
    Suite*.\
-   **Salary Tier:** High-paying positions require advanced skills like
    *AI Ethics* and *Cloud Architecture*; lower-paying roles rely on *MS
    Office* and *Basic SQL*.

## Conclusion {.scrollable}

::: {style="font-size:0.8em; line-height:1.2; width:90vw; margin:auto; max-height:80vh; overflow-y:auto;"}
### Key Findings

-   **Remote roles favor technical skills:** Python, Cloud Architecture,
    Cybersecurity and Machine Learning top the list for fully remote
    positions.\
-   **In-person roles emphasize soft skills:** Communication, Teamwork
    and Project Management remain critical where on-site collaboration
    is required.\
-   **Industry and pay tiers diverge:** Tech industries and high-salary
    jobs demand advanced AI-related expertise; lower-paid and non-tech
    roles rely on foundational tools like MS Office and basic SQL.

### Implications for Businesses

-   **Targeted upskilling:** Invest in AI/data training for remote teams
    and soft-skill workshops for in-house staff.\
-   **Strategic hiring:** Tailor job postings—highlight technical skills
    for remote, human-centric for on-site.\
-   **Flexible work models:** Leverage remote policies to access niche
    talent while preserving on-site culture.\
-   **Workforce planning:** Use these insights to forecast talent needs,
    align L&D budgets, and shape long-term recruitment strategies.\
:::
