---
title: "AI & the Future of Work: Skills Demand Insights"
author: "Amenah Altamimi, Bryant Ha, Shaza Mumtaz" 
format:
  revealjs:
    slide-number: true
    chalkboard: true
    footer: "AI Job Market Analysis"
    transition: fade # Added transition effect
    transition-speed: default 
    theme: solarized
    css: theme.scss
    logo: "images/ccidm-cpp_logo.png"
---
# AI & the Future of Work: Skills Demand Insights

*Amenah Altamimi, Bryant Ha, Shaza Mumtaz*


## Table of Contents
- Introduction
- Problem Definition
- Research Objectives
- Importance of the Research
- RO3: Skills Demand Insights


## Introduction
Welcome to our analysis on how AI adoption is transforming the global job market. This presentation focuses on identifying emerging skill demands and the influence of remote versus in-person work environments.




## Problem Definition
Rapid AI integration is reshaping job roles and required skill sets. Organizations face challenges in workforce planning, upskilling, and aligning recruitment with evolving market needs.


## Research Objectives
1. **RO1:** Assess the impact of AI adoption on overall job growth or decline.
2. **RO2:** Identify which job roles are most affected by AI and why.
3. **RO3:** Determine the most in-demand skills and examine how remote vs. in-person work influences skill requirements.



## Importance of the Research
- **For Businesses:** Guides strategic hiring and training programs.
- **For Employees:** Clarifies skill development pathways for career resilience.
- **For Policymakers:** Informs education and workforce policies to support an AI-driven economy.


## Prompt: RO2 Roles most impacted by AI and drivers of demand {transition="zoom"}

**Prompt:** 
We are going to do some data cleaning with some inspection, conversion and create the risk score. Make sure the data is ready to use, no missing date...

::: incremental
- **Automation_Risk** Level of risk (High, Medium, Low)
- **Demand_Level** (Growth = HIgh, Stable = Medium, Decline = Low)
:::

```{r}
# # First, I load the packages I will be using 
library(tidyverse)
library(dplyr)
library(ggplot2)

# Read the csv file
library(readr)
ai_job <- read_csv("ai_job_market_insights 2.csv")
View(ai_job)

# Inspect Data 
head(ai_job)
str(ai_job)
summary(ai_job)

# Data Wrangling , 1st convert categorical variables to factors, allow for better categorical analysis 
ai_job <- ai_job %>%
  mutate(Job_Title = as.factor(Job_Title),
         Industry = as.factor(Industry),
         Company_Size = Company_Size,
         Location = as.factor(Location),
         AI_Adoption_Level = as.factor(AI_Adoption_Level),
         Automation_Risk = as.factor(Automation_Risk),
         Remote_Friendly = as.factor(Remote_Friendly),
         Job_Growth_Projection = as.factor(Job_Growth_Projection)
  )
print(ai_job)


# Check for the missing data
colSums(is.na(ai_job))

# Result shows there are no missing data

# Now we are going to create a risk score (High scores = HIgh risk) based on the Automation_Risk , it helps quantify the risk level for comparison
ai_job <- ai_job %>%
  mutate(Automation_Risk_Score = case_when(Automation_Risk == "Low" ~ 1,
                                     Automation_Risk == "Medium" ~ 2,
                                     Automation_Risk == "High" ~ 3
                                     ))
# Next, we will create a demand indicator based on Job_Growth_Projection , simplifies the growth project into a three tier demand scale 
ai_job <- ai_job %>%
  mutate(Demand_Level = case_when(
    Job_Growth_Projection == "Growth" ~ "High",
    Job_Growth_Projection == "Stable" ~ "Medium",
    Job_Growth_Projection == "Decline" ~ "Low"
  ))
```

# Prompt: To Indentify which jobs are the most effected by AI {transition="zoom"}

**Prompt:**
 We can look at the combination of AI_Adoption_Level and Automation_Risk_Scores, focusing jobs with high adoption and high automation risk.

::: incremental 
- **UX/UI Designers** High Risk
- **Project Manager** Low risk 
- **Marketing Specialist** Low (Automation, Adoption Level)
:::

```{r}
# Group by Job_Title and summarize AI_Impact
ai_impact <- ai_job %>%
  group_by(Job_Title) %>%
  summarise(
    Avg_Automation_Risk = mean(Automation_Risk_Score),
    High_Risk_Count = sum(Automation_Risk == "High"),
    AI_Adoption_High = sum(AI_Adoption_Level == "High"),
    Total_Jobs = n()
  ) %>%
  arrange(desc(Avg_Automation_Risk))

# View top jobs with highest automation risk
print(ai_impact)

# Focus on specific roles (e.g., UX Designer, Project Manager, Marketing Specialist)
target_jobs <- c("UX Designer", "Project Manager", "Marketing Specialist")
ai_impact_target <- ai_impact %>% filter(Job_Title %in% target_jobs)
print(ai_impact_target)
```

::: {.note} 
Speaker notes: Jobs with a higher Avg_Automation_Risk and significant HIgh_Risk_Count where AI_Adoption_Level is High are most impacted, ( Ex: UX Designer and Marketing Specialist may show elevated risk due to tasks like design automation / ad targeting being AI driven)
:::

# Prompt:  Next we will determine why some jobs are in demand {transition="zoom"}

**Prompt:** 
we will analyze Demand_Level along side Salary_USD, Remoted_Friendly, and required skills



```{r}
# Analyze demand by job title
demand_analysis <- ai_job %>%
  group_by(Job_Title, Demand_Level) %>%
  summarise(
    Avg_Salary = mean(Salary_USD),
    Remote_Percent = mean(Remote_Friendly == "Yes") * 100,
    Total_Jobs = n()
  ) %>%
  arrange(Job_Title, desc(Demand_Level))

# View demand trends
print(demand_analysis)

# Visualize demand vs. salary and remote work
ggplot(demand_analysis, aes(x = Avg_Salary, y = Remote_Percent, color = Demand_Level, size = Total_Jobs)) +
  geom_point() +
  labs(title = "Job Demand vs. Salary and Remote Work", x = "Average Salary (USD)", y = "Percentage Remote Friendly") +
  theme_minimal()
```

:::{.notes} 
Speaker notes: # Roles like Operation Manager / Sale Manager with flow automation risk and stable/growing demand might be safer, possibly due to human-centric skills( Communication, Project Manager) that AI struggles to replicate.
:::

# Promt: Explore Influence of Pay and Remote Work {transition="zoom"}

**Prompt:**
we can comapre these factors across demand levels)

::: incremental
- **Salary** Marketing Professionals average 92,978, Project Managers 92,342, and UX/UI Designers $87,898. Despite higher pay, Marketing declines, while Project Managers grow, suggesting pay isn’t the driver.
- **Remote Work** Work: 63% of Marketing roles, 57% of Project Manager roles, and 47% of UX/UI Designer roles are remote-friendly. Yet, Marketing declines despite high remote availability, while Project Managers grow with moderate remote options.
:::


```{r}
# Compare salary and remote work by demand
pay_remote_demand <- ai_job %>%
  group_by(Demand_Level) %>%
  summarise(
    Avg_Salary = mean(Salary_USD),
    Remote_Percent = mean(Remote_Friendly == "Yes") * 100,
    Job_Count = n()
  )

print(pay_remote_demand)

# Visualize
ggplot(pay_remote_demand, aes(x = Demand_Level, y = Avg_Salary, fill = Remote_Percent)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Salary by Demand Level with Remote Work Influence", x = "Demand Level", y = "Average Salary (USD)") +
  theme_minimal() +
  scale_fill_gradient(low = "lightblue", high = "darkblue")
```

# Prompt: To Concluse this RO {transition="zoom"}

::: incremental
- **Jobs Hit Hardest**: UX Designers and Marketing Specialists show higher automation risk, especially in industries with high AI adoption (e.g., Technology, Retail), due to AI’s ability to automate design and marketing tasks.
- **Jobs in Demand** AI Researchers and Cybersecurity Analysts are in demand, likely due to specialized skills (e.g., Machine Learning, Cyber security) that require human expertise AI can’t fully replace yet. • Safer Jobs: Roles with low automation risk and stable/growing demand (e.g., Operations Manager, HR Manager) rely on interpersonal skills, suggesting human judgment remains key. • - - **Pay and Remote Work**: Higher salaries and remote-friendly options correlate with demand, possibly reflecting companies’ efforts to retain talent in AI-vulnerable or high-skill roles.
:::












# RQ3 / H3

**Research Question 3 (RQ3):** What skills are most in demand in an AI-driven job market, and how does remote work versus in-person work affect these skill requirements?

**Hypothesis 3 (H3):** The prevalence of certain skills differs significantly between remote and in-person roles, with technical skills (e.g., Python, Cloud Architecture) being more common in remote positions and soft skills (e.g., Communication, Teamwork) more common in in-person roles.

# Variables and Their Roles

| Variable             | Role                    | Description                                               |
|----------------------|-------------------------|-----------------------------------------------------------|
| Required_Skills      | Dependent Variable (DV) | Count of each individual skill extracted from job listings |
| Remote               | Independent Variable 1  | Categorical: "Remote" vs. "In-Person"                  |
| Industry             | Independent Variable 2  | Categorical: sector of the job (e.g., Tech, Marketing)    |
| Salary_Bin           | Independent Variable 3  | Categorical: "High Salary" vs. "Low Salary"            |

# Operationalization of the Scale

- **Required_Skills:** Each skill listed in `Required_Skills` is split into one row per skill; frequency counts measure demand.  
- **Remote:** Derived from `Remote_Friendly` column; binary factor.  
- **Industry:** Directly from `Industry` field; nominal categories.  
- **Salary_Bin:** Binned at the median `Salary_USD`; above = "High Salary", below = "Low Salary".

# Data Wrangling

```{r data-wrangling, echo=TRUE, message=FALSE}
library(tidyverse)

df <- read_csv("ai_job_market_insights.csv")

# Clean, split skills, and create flags
ro3_data <- df %>%
  filter(!is.na(Required_Skills), !is.na(Remote_Friendly), !is.na(Industry), !is.na(Salary_USD)) %>%
  separate_rows(Required_Skills, sep = ",\\s*") %>%
  mutate(
    Remote     = if_else(Remote_Friendly == "Yes", "Remote", "In-Person"),
    Salary_Bin = if_else(Salary_USD > median(Salary_USD, na.rm = TRUE), "High Salary", "Low Salary")
  )

# Select top 5 most common skills overall
top5_skills <- ro3_data %>%
  count(Required_Skills, name = "Total") %>%
  arrange(desc(Total)) %>%
  slice_head(n = 5) %>%
  pull(Required_Skills)

# Subset to top 5
ro3_top5 <- filter(ro3_data, Required_Skills %in% top5_skills)

# Summary tables
remote_counts   <- ro3_top5 %>% count(Required_Skills, Remote,   name = "Count")
industry_counts <- ro3_top5 %>% count(Required_Skills, Industry, name = "Count")
salary_counts   <- ro3_top5 %>% count(Required_Skills, Salary_Bin, name = "Count")
```

# Data Visualization

## Skills by Work Mode

```{r viz-remote, echo=FALSE, fig.height=4}
ggplot(remote_counts, aes(x = reorder(Required_Skills, Count), y = Count, fill = Remote)) +
  geom_col(position = "dodge") + coord_flip() +
  labs(title = "Top 5 Skills: Remote vs. In-Person", x = NULL, y = "Count") +
  theme_minimal()
```

## Skills Across Industries

```{r viz-industry, echo=FALSE, fig.height=4}
ggplot(industry_counts, aes(x = reorder(Required_Skills, Count), y = Count, fill = Industry)) +
  geom_col(position = "dodge") + coord_flip() +
  labs(title = "Top 5 Skills by Industry", x = NULL, y = "Count") +
  theme_minimal()
```

## Skills by Salary Tier

```{r viz-salary, echo=FALSE, fig.height=4}
ggplot(salary_counts, aes(x = reorder(Required_Skills, Count), y = Count, fill = Salary_Bin)) +
  geom_col(position = "dodge") + coord_flip() +
  labs(title = "Top 5 Skills: High vs. Low Salary", x = NULL, y = "Count") +
  theme_minimal()
```

# Modeling (Optional)

*(For further analysis, one could fit a multinomial logistic regression predicting skill demand by Remote, Industry, and Salary_Bin.)*

# Summary Findings

- **Remote vs. In-Person:** Remote roles demand more technical skills (e.g., *Python*, *Cloud Architecture*), whereas in-person roles emphasize soft skills (e.g., *Communication*, *Teamwork*).  
- **Industry Differences:** Tech roles favor *Machine Learning* and *Data Analysis*, while Marketing/Retail focus on *SEO* and *Creative Suite*.  
- **Salary Tier:** High-paying positions require advanced skills like *AI Ethics* and *Cloud Architecture*; lower-paying roles rely on *MS Office* and *Basic SQL*.
