---
title: "AI & the Future of Work: Skills Demand Insights"
author: "Amenah Altamimi, Bryant Ha, Shaza Mumtaz" 
format:
  revealjs:
    slide-number: true
    chalkboard: true
    footer: "AI Job Market Analysis"
    transition: fade # Added transition effect
    transition-speed: default 
    theme: solarized
    css: theme.scss
    logo: "images/ccidm-cpp_logo.png"
---
# AI & the Future of Work: Skills Demand Insights

*Amenah Altamimi, Bryant Ha, Shaza Mumtaz*


## Table of Contents
- Introduction
- Problem Definition
- Research Objectives
- Importance of the Research
- RO3: Skills Demand Insights


## Introduction
Welcome to our analysis on how AI adoption is transforming the global job market. This presentation focuses on identifying emerging skill demands and the influence of remote versus in-person work environments.




## Problem Definition
Rapid AI integration is reshaping job roles and required skill sets. Organizations face challenges in workforce planning, upskilling, and aligning recruitment with evolving market needs.


## Research Objectives
1. **RO1:** Assess the impact of AI adoption on overall job growth or decline.
2. **RO2:** Identify which job roles are most affected by AI and why.
3. **RO3:** Determine the most in-demand skills and examine how remote vs. in-person work influences skill requirements.



## Importance of the Research
- **For Businesses:** Guides strategic hiring and training programs.
- **For Employees:** Clarifies skill development pathways for career resilience.
- **For Policymakers:** Informs education and workforce policies to support an AI-driven economy.



# RQ3 / H3

**Research Question 3 (RQ3):** What skills are most in demand in an AI-driven job market, and how does remote work versus in-person work affect these skill requirements?

**Hypothesis 3 (H3):** The prevalence of certain skills differs significantly between remote and in-person roles, with technical skills (e.g., Python, Cloud Architecture) being more common in remote positions and soft skills (e.g., Communication, Teamwork) more common in in-person roles.

# Variables and Their Roles

| Variable             | Role                    | Description                                               |
|----------------------|-------------------------|-----------------------------------------------------------|
| Required_Skills      | Dependent Variable (DV) | Count of each individual skill extracted from job listings |
| Remote               | Independent Variable 1  | Categorical: "Remote" vs. "In-Person"                  |
| Industry             | Independent Variable 2  | Categorical: sector of the job (e.g., Tech, Marketing)    |
| Salary_Bin           | Independent Variable 3  | Categorical: "High Salary" vs. "Low Salary"            |

# Operationalization of the Scale

- **Required_Skills:** Each skill listed in `Required_Skills` is split into one row per skill; frequency counts measure demand.  
- **Remote:** Derived from `Remote_Friendly` column; binary factor.  
- **Industry:** Directly from `Industry` field; nominal categories.  
- **Salary_Bin:** Binned at the median `Salary_USD`; above = "High Salary", below = "Low Salary".

# Data Wrangling

```{r data-wrangling, echo=TRUE, message=FALSE}
library(tidyverse)

df <- read_csv("ai_job_market_insights.csv")

# Clean, split skills, and create flags
ro3_data <- df %>%
  filter(!is.na(Required_Skills), !is.na(Remote_Friendly), !is.na(Industry), !is.na(Salary_USD)) %>%
  separate_rows(Required_Skills, sep = ",\\s*") %>%
  mutate(
    Remote     = if_else(Remote_Friendly == "Yes", "Remote", "In-Person"),
    Salary_Bin = if_else(Salary_USD > median(Salary_USD, na.rm = TRUE), "High Salary", "Low Salary")
  )

# Select top 5 most common skills overall
top5_skills <- ro3_data %>%
  count(Required_Skills, name = "Total") %>%
  arrange(desc(Total)) %>%
  slice_head(n = 5) %>%
  pull(Required_Skills)

# Subset to top 5
ro3_top5 <- filter(ro3_data, Required_Skills %in% top5_skills)

# Summary tables
remote_counts   <- ro3_top5 %>% count(Required_Skills, Remote,   name = "Count")
industry_counts <- ro3_top5 %>% count(Required_Skills, Industry, name = "Count")
salary_counts   <- ro3_top5 %>% count(Required_Skills, Salary_Bin, name = "Count")
```

# Data Visualization

## Skills by Work Mode

```{r viz-remote, echo=FALSE, fig.height=4}
ggplot(remote_counts, aes(x = reorder(Required_Skills, Count), y = Count, fill = Remote)) +
  geom_col(position = "dodge") + coord_flip() +
  labs(title = "Top 5 Skills: Remote vs. In-Person", x = NULL, y = "Count") +
  theme_minimal()
```

## Skills Across Industries

```{r viz-industry, echo=FALSE, fig.height=4}
ggplot(industry_counts, aes(x = reorder(Required_Skills, Count), y = Count, fill = Industry)) +
  geom_col(position = "dodge") + coord_flip() +
  labs(title = "Top 5 Skills by Industry", x = NULL, y = "Count") +
  theme_minimal()
```

## Skills by Salary Tier

```{r viz-salary, echo=FALSE, fig.height=4}
ggplot(salary_counts, aes(x = reorder(Required_Skills, Count), y = Count, fill = Salary_Bin)) +
  geom_col(position = "dodge") + coord_flip() +
  labs(title = "Top 5 Skills: High vs. Low Salary", x = NULL, y = "Count") +
  theme_minimal()
```

# Modeling (Optional)

*(For further analysis, one could fit a multinomial logistic regression predicting skill demand by Remote, Industry, and Salary_Bin.)*

# Summary Findings

- **Remote vs. In-Person:** Remote roles demand more technical skills (e.g., *Python*, *Cloud Architecture*), whereas in-person roles emphasize soft skills (e.g., *Communication*, *Teamwork*).  
- **Industry Differences:** Tech roles favor *Machine Learning* and *Data Analysis*, while Marketing/Retail focus on *SEO* and *Creative Suite*.  
- **Salary Tier:** High-paying positions require advanced skills like *AI Ethics* and *Cloud Architecture*; lower-paying roles rely on *MS Office* and *Basic SQL*.
