---
title: "AI Job Market Analysis Code"
format: html
execute:
  warning: false
  message: false
---

# Introduction

This report analyzes the relationship between AI adoption levels and job growth projections across industries, using data from the `ai_job_market_insights.csv` dataset. The analysis includes descriptive statistics, visualizations, and statistical tests to explore trends and patterns.

# Setup

Load required R libraries for data manipulation, visualization, and statistical analysis.

```{r setup, message=FALSE}
library(tidyverse)
library(ggthemes)
library(knitr)
library(MASS)
```

# 1. Data Preparation

Load and clean the dataset, creating a processed dataset with factored AI adoption levels and a numeric growth score.

```{r data-prep}
# Load and clean data
df <- read_csv("ai_job_market_insights.csv")

ro1_data <- df %>%
  filter(!is.na(AI_Adoption_Level), !is.na(Job_Growth_Projection)) %>%
  mutate(
    AI_Adoption = factor(AI_Adoption_Level, 
                         levels = c("Low", "Medium", "High"),
                         ordered = TRUE),
    Growth_Score = case_when(
      Job_Growth_Projection == "Decline" ~ -1,
      Job_Growth_Projection == "Stable" ~ 0,
      Job_Growth_Projection == "Growth" ~ 1
    )
  )
```

# 2. Descriptive Statistics

Summarize job counts by AI adoption level and job growth projection.

```{r desc-stats}
desc_stats <- ro1_data %>%
  count(AI_Adoption, Job_Growth_Projection) %>%
  pivot_wider(names_from = Job_Growth_Projection, values_from = n)

kable(desc_stats, caption = "Job Counts by AI Adoption Level")
```

# 3. Visualizations

## Growth Proportion Plot

Visualize the proportion of job growth statuses by AI adoption level.

```{r growth-proportion, fig.width=8, fig.height=6}
ggplot(ro1_data, aes(x = AI_Adoption, fill = Job_Growth_Projection)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Job Growth Proportion by AI Adoption Level",
    x = "AI Adoption Level",
    y = "Proportion",
    fill = "Growth Status"
  ) +
  theme_minimal()
```

## Growth Score Boxplot

Display the distribution of growth scores across AI adoption levels.

```{r growth-boxplot, fig.width=8, fig.height=6}
ggplot(ro1_data, aes(x = AI_Adoption, y = Growth_Score, fill = AI_Adoption)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Job Growth Score by AI Adoption Level",
    x = "AI Adoption Level",
    y = "Growth Score (Decline=-1, Stable=0, Growth=1)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Industry Facet Plot

Show job growth by AI adoption level across different industries.

```{r industry-facet, fig.width=8, fig.height=6}
ggplot(ro1_data, aes(x = AI_Adoption, fill = Job_Growth_Projection)) +
  geom_bar(position = "dodge") +
  facet_wrap(~Industry, ncol = 2) +
  labs(
    title = "Job Growth by AI Adoption Across Industries",
    x = "AI Adoption Level",
    y = "Count",
    fill = "Growth Status"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# 4. Statistical Analysis

## Chi-Squared Test

Test for association between AI adoption level and job growth projection.

```{r chi-test}
chi_test <- chisq.test(
  table(ro1_data$AI_Adoption_Level, ro1_data$Job_Growth_Projection))
print(chi_test)
```

## Ordinal Regression

Fit an ordinal regression model to predict job growth projection based on AI adoption level (if sufficient data is available).

```{r ordinal-regression}
if (nrow(ro1_data) > 50) {
  growth_ordinal <- polr(
    factor(Job_Growth_Projection, 
           levels = c("Decline", "Stable", "Growth")) ~ AI_Adoption,
    data = ro1_data,
    Hess = TRUE
  )
  print(summary(growth_ordinal))
} else {
  cat("Insufficient data for ordinal regression (n <= 50).\n")
}
```

# 5. Key Metrics

Summarize average growth scores by AI adoption level and the proportion of growing jobs by industry and AI adoption.

```{r key-metrics}
cat("\nKey Metrics:\n")
cat("Average growth score by AI level:\n")
ro1_data %>%
  group_by(AI_Adoption) %>%
  summarise(Mean_Growth = mean(Growth_Score)) %>%
  print()

cat("\nIndustry breakdown:\n")
ro1_data %>%
  group_by(Industry, AI_Adoption) %>%
  summarise(Prop_Growth = mean(Growth_Score > 0), .groups = "drop") %>%
  arrange(Industry, AI_Adoption) %>%
  print(n = Inf)
```

# 6. Export Results

Save the processed dataset and visualization outputs.

```{r export}
write_csv(ro1_data, "ro1_processed_data.csv")
ggsave("growth_proportion.png", width = 8, height = 6)
ggsave("growth_score_distribution.png", width = 8, height = 6)
```

# Conclusion

The analysis suggests that AI_Adoption (independent variable) has a nuanced but not statistically significant impact on Job_Growth_Projection (dependent variable), with a chi-squared p-value of 0.667 and weak ordinal regression coefficients. Growth scores remain balanced across AI adoption levels (Low: 0.052, Medium: 0, High: -0.061), indicating stable job markets. Industries like Transportation (0.636 at medium AI adoption) and Finance (0.579 at low AI adoption) show promising growth, highlighting opportunities for job expansion in AI-integrated sectors. Further exploration of industry-specific factors could unlock additional growth potential.
